---
title: "STA5076Z Lecture 12 -- Gradient boosting"
format: 
  html:
    fig-width: 6
    fig-height: 6
    code-fold: show
    code-summary: "Show/Hide Code"
    code-overflow: wrap
    code-tools: true
execute:
  warning: false
bibliography: references.bib
---

```{css}
/*| echo: false*/
p {
text-align: justify;
}

table {
width: auto !important;
margin-left: auto;
margin-right: auto;
}
```
### Boosting

We begin by initiating the algorithm discussed in the lecture from scratch and applying it to a dummy data set.

```{r}
library(rpart)
set.seed(5076)
n <- 100
x1 <- round(runif(n, 0, 10), 0)
x2 <- round(runif(n, 0, 10), 0)
y <- round(10 + 3*x1 + x2 + rnorm(n, 0, 10))
dat <- data.frame(cbind(y, x1, x2))

# Initiate algorithm "by hand"
gb.manual <- function(lambda, d, ntrees){
  train_mse <<- c()                     #To keep track of training error
  resids <<- c()                        #To keep track of residuals
  rhats <<- c()                         #To keep track of tree predictions
  
  yhat_b <- mean(y)                     #First initialisation of prediction -> step 1
  r_b <- y - yhat_b                     #First initialisation of residuals  -> step 1
  
  for(b in 1:ntrees){
    ctrl <- rpart.control(maxdepth = d, cp = 0.001)  
    t <- rpart(r_b ~ x1 + x2, 
               control = ctrl)          #Fit tree to residuals              -> step 2a
    rhat_b <- predict(t, newdata = dat) #Predicted values of tree b         -> step 2b
    yhat_b <- yhat_b + lambda*rhat_b    #Updated prediction                 -> step 2b
    r_b <- y - yhat_b                   #Updated residuals                  -> step 2c
    # r_b <- r_b - lambda*rhat_b        #Could also write it like this, same thing
    
    # Save all the training mses, residuals, and tree predictions
    train_mse <<- c(train_mse, mean((y - yhat_b)^2)) #Double arrow saves object outside function
    resids <<- cbind(resids, r_b)
    rhats <<- cbind(rhats, rhat_b)
  }
  return(mean(y) + lambda*apply(rhats, 1, sum)) #Output final predictions   -> step 3
  # Could also return y - resids[, ntrees]
  # Or just yhat_b
}

#Run algorithm
yhat_manual <- gb.manual(lambda = 0.01, d = 1, ntrees = 1000)

#Final model training error
(gbm_manual_train_mse <- mean((y - yhat_manual)^2)) # Or just train_mse[ntrees]
```

In order to validate this procedure, let us compare the results with those obtained from using the `gbm` package.[^gbm-note]

[^gbm-note]: The `gbm` package is no longer in development and has been succeeded by [gbm3](https://github.com/gbm-developers/gbm3). Although the latter was removed from the CRAN repository in February 2024, it can be installed via github.

```{r warning=F, message=F}
library(gbm)
gbm_lib <- gbm(y ~ ., data = dat, 
               distribution = 'gaussian', #Applies squared error loss 
               n.trees = 1000,
               interaction.depth = 1,
               shrinkage = 0.01,
               bag.fraction = 1)          #Subsamples observations by default (0.5)

yhat_gbm <- gbm_lib$fit
(mse_gbm <- mean((y - yhat_gbm)^2))
```

We see that the training MSEs match exactly. As a final sanity check, let's compare each tree's training error between the manual implementation and the package's.

```{r}
all(round(gbm_lib$train.error, 9) == round(train_mse, 9))
```

Note that some rounding issue creeps in after 9 decimal places (exacerbated by larger trees).

### Example -- California housing

Since this is our final analysis of this dataset, we will apply some **feature engineering**, a process of adapting/transforming/combining the existing information into new features to improve model performance.

Instead of using the total number of bedrooms, bathrooms, and people per neighbourhood, let's use the number of households to convert these into per-house rates. We will also apply a 80/20 train-test split, here showing how a `caret` function can be used to do so.

```{r}
library(dplyr)
library(caret)

calif <- read.csv('data/calif.csv')
calif$HousePrice <- log(calif$HousePrice)

# Create calif feature engineered
calif_fe <- calif |> mutate(AveRooms = Rooms/Households,
                            AveBedrms = Bedrooms/Households,
                            AveOccup = Population/Households) |>
  select(-Rooms, -Bedrooms, -Households)

## Using caret for train/test split:
set.seed(4026)
train_index <- createDataPartition(y = calif_fe$HousePrice, p = 0.8, list = FALSE)
calif_fe_train <- calif_fe[train_index,]
calif_fe_test <- calif_fe[-train_index,]
```

As some benchmark for comparison -- and to see whether the engineered features are useful -- we will first fit a single regression tree.

```{r}
#| label: fig-calif_fe_cv
#| fig-cap: Cross-validated pruning results from a tree fitted to the feature-engineered Cailfornia housing training set.

library(tree)

# Slightly larger tree
calif_fe_tree <- tree(HousePrice ~ ., data = calif_fe_train, 
                      control = tree.control(nobs=nrow(calif), 
                                             mindev = 0.003))

# Tune
set.seed(1)
calif_fe_cv <- cv.tree(calif_fe_tree)
calif_fe_cv$k[1] <- 0

plot(calif_fe_cv$size, calif_fe_cv$dev, type = 'o', 
     pch = 16, col = 'navy', lwd = 2,
     xlab = 'Number of terminal nodes', ylab = 'CV RSS')
axis(side = 1, at = 1:max(calif_fe_cv$size), labels = 1:max(calif_fe_cv$size))
alpha <- round(calif_fe_cv$k)
axis(3, at = calif_fe_cv$size, lab = alpha)
mtext(expression(alpha), 3, line = 2.5, cex = 1.5)
```

The CV RSS first reaches a minimum at 12 terminal nodes, hence we will prune it down to that size.

```{r}
#| label: fig-calif_fe_pruned
#| fig-cap: Pruning tree fitted to the feature-engineered Cailfornia housing training set.

calif_fe_pruned <- prune.tree(calif_fe_tree, best = 12)
plot(calif_fe_pruned)
text(calif_fe_pruned)
```

Here we see that the average number of rooms and average occupancy were indeed useful in this model (compare to our pruned tree from the regression trees lecture, which split almost exclusively on income and location). The CV RMSE is `r round(sqrt(calif_fe_cv$dev[which(calif_fe_cv$size==12)]/nrow(calif_fe_train)), 3)`.

Next, we fit a gradient boosted trees model to the training set via the `gbm` package, using `caret` to apply hyperparameter tuning according to 10-fold CV (which can be parallelised). Note that for a problem of this size, the gbm model does take a while to run. Therefore, the code run below has been commented out after running, with the results saved and loaded back in for quick reproducibility.

```{r}
# set.seed(4026)
# ctrl <- trainControl(method = 'cv', number = 10, verboseIter = T)
# 
# calif_gbm_grid <- expand.grid(n.trees = c(1000, 10000, 50000),
#                               interaction.depth = c(1, 2, 5),
#                               shrinkage = c(0.01, 0.001),
#                               n.minobsinnode = 10)
# 
# calif_gbm_gridsearch <- train(HousePrice ~ ., data = calif_fe_train,
#                               method = 'gbm',
#                               distribution = 'gaussian',
#                               trControl = ctrl,
#                               verbose = T,
#                               tuneGrid = calif_gbm_grid)
# save(calif_gbm_gridsearch, file = 'data/gbm_calif_50k.Rdata')
load('data/gbm_calif_50k.Rdata')
```

Although we only touch on the subject in this course, the hyperparameter tuning procedure can be an entire topic on its own. One alternative to the grid search applied above -- whereby all combinations of **fixed** sets of hyperparameter values are used -- is to apply a random search, which introduces some variation in the hyperparameters. @bergstra2012random initially provided a strong argument for this approach within a neural network context. Further exploration of these topics is left for the reader.

Looking at the CV results in @fig-calif_gbm_hps, we see that higher values of $d$ in fact yielded better results. As expected, the higher learning rate reduced the CV error quicker for the same number of trees.

```{r}
#| label: fig-calif_gbm_hps
#| fig-cap: Hyperparameter combinations results for a gradient boosted trees model fitted to the feature-engineered California housing dataset.

plot(calif_gbm_gridsearch)
```

However, it is important to note that for this particular combination of number of trees it is difficult to determine whether the error is still decreasing at 50,000 trees. To ascertain this, we will use a large number of trees and see where the CV error reaches a minimum, as shown in @fig-calif_gbm_cv.

```{r}
#| label: fig-calif_gbm_cv
#| fig-cap: Cross-validation results for the gradient boosted trees model with $\lambda=0.01$ and $d=5$ fitted to the feature-engineered California housing dataset.
options(scipen = 999)

# calif_fe_gbm <-   gbm(HousePrice ~ ., data = calif_fe_train, 
#                       distribution = 'gaussian', 
#                       n.trees = 100000, 
#                       interaction.depth = 5, 
#                       shrinkage = 0.01, 
#                       bag.fraction = 1, 
#                       cv.folds = 10,  #built-in CV
#                       n.cores = 10,   #which can be parallelised
#                       verbose = T)
# save(calif_fe_gbm, file = 'data/gbm_calif_100k.Rdata')
load('data/gbm_calif_100k.Rdata')
d <- gbm.perf(calif_fe_gbm)
legend('topright', c('CV error', 'Training error'), col = c('green', 'black'), lty = 1)
```

It turns out the CV error actually started increasing after `r d` trees. However, it is possible that one of the other hyperparameter combinations (different $\lambda$ and $d$) would have yielded a lower CV if allowed to run for more than 50,000 trees. This example serves to illustrate that this style of hyperparameter tuning can be tedious, time-consuming, and a trial-and-error process. Whilst more sophisticated tuning techniques exist, for example Bayesian optimisation, each has their own compromises and drawbacks. These are beyond the scope of this course.

The **variable importance** plots (this time using `vi_permute()`) are shown in Figure \@ref(fig:calif-gbm-vi):

```{r calif-gbm-vi, fig.align='center', fig.cap='Variable importance for the tuned gbm model fitted to the feature-engineered California housing dataset.'}
library(vip)
library(patchwork)

set.seed(1)

x_train <- select(calif_fe_train, -HousePrice)
y_train <- calif_fe_train$HousePrice

cgbm_pred_wrapper <- function(object, newdata) {
  predict(object, newdata = newdata, n.trees = d)  # d = chosen number of trees above
}

# cgbm_per_imp <- vi_permute(
#   object        = calif_fe_gbm,
#   feature_names = colnames(x_train),
#   train         = x_train,
#   target        = y_train,
#   metric        = 'rmse',
#   pred_wrapper  = cgbm_pred_wrapper,
#   nsim          = 5
# )
# save(cgbm_per_imp, file = 'data/calif_fe_perm_varimp.RData')
load('data/calif_fe_perm_varimp.RData')

cgbm_per_p <- vip(cgbm_per_imp) +
  labs(
    title = 'Permutation Variable Importance',
    x = 'Increase in Prediction Error',
    y = NULL
  ) +
  theme_minimal()

cgbm_imp_p <- vip(calif_fe_gbm) +
  labs(
    title = 'Impurity Variable Importance',
    x = 'Decrease in Splitting Criterion',
    y = NULL
  ) +
  theme_minimal()

cgbm_per_p + cgbm_imp_p
```

We see that according to impurity-based importance, `Income` appears dominant, implying that it is highly effective at reducing training RSS and is frequently used in early splits. Permutation-based importance, on the other hand, reflects the extent to which the model’s predictive performance depends _uniquely_ on a variable. Since `Income` is strongly correlated with the spatial predictors `Latitude` and `Longitude`, permuting it does not degrade predictive accuracy as dramatically as its split-based contribution might suggest. This highlights that impurity importance measures how much a variable is used in constructing the model, whereas permutation importance measures how much the model relies on that variable for prediction.

It's also worth noting that two of our engineered features -- average occupancy and average rooms -- were somewhat influential according to both measures.

Although variable importance plots tell us which features are most important for predicting the response, they do not tell us **how** the predictors influence the predictions. For this, we can make use of partial dependence plots.

### Partial dependence plots

Visualisation of the relationship between the predictors and the predicted outcome is limited to low-dimensional views. Although these "slices" of information can seldom provide a comprehensive depiction of the approximation, it can still provide a useful indication of how certain features influenced the predictions on average, especially when their interaction with other variables is limited.

The `gbm` package again allows us to easily produce the partial dependence plots. @fig-calif_pdp shows 1-dimensional partial dependence plots for the previously fitted model on the (feature-engineered) California dataset for four of the (non-location) features.

```{r}
#| label: fig-calif_pdp
#| fig-cap: Partial dependence plots for the gbm model fitted to the feature-engineered California housing dataset.

library(gridExtra)
library(grid)

ylims <- c(11.5,12.8)
p1 <- plot.gbm(calif_fe_gbm, 'Income', d, ylim = ylims, ylab = '')
p2 <- plot.gbm(calif_fe_gbm, 'AveRooms', d, ylim = ylims, ylab = '', xlab = 'Average Rooms')
p3 <- plot.gbm(calif_fe_gbm, 'AveOccup', d, ylim = ylims, ylab = '', xlab = 'Average Occupancy')
p4 <- plot.gbm(calif_fe_gbm, 'HouseAge', d, ylim = ylims, ylab = '', xlab = 'House Age')

grid.arrange(p1, p2, p3, p4, ncol = 2)
grid.text('Partial Predicted Log House Price', x = 0, vjust = 1, rot = 90)
```

Looking at the average occupancy and average rooms, something is not right -- an average occupancy of more than 1000 does not make sense, nor does more than 100 rooms on average! It is most likely an errant observation(s) with too few households.

```{r}
calif[which.max(calif_fe$AveOccup),]
calif[which.max(calif_fe$AveRooms),]
```

Indeed, six households for a population of 7460 is nonsensical. 1561 rooms for 11 households and a population of 30 is questionable, given some of the areas filled with mansions one finds in California. This goes to show how important an exploratory data analysis and cleaning are at the start of an analysis! These are just the observations yielding the maximum average occupancy and average rooms values, one could explore their distributions further.

It would be best to rerun the analysis with the cleaned data, which is left as a homework exercise. For now, we will just cut off these outliers, for which we can use a package offering more options for partial dependence plots. `pdp` is a popular package for this.

```{r}
#| label: fig-calif_pdp2
#| fig-cap: Partial dependence plots for some of the features in the gbm model fitted to the feature-engineered California housing dataset. Some of the x-axes have been rescaled.

library(pdp)

p1 <- partial(calif_fe_gbm, 'Income',   n.trees = d)
p2 <- partial(calif_fe_gbm, 'AveRooms', n.trees = d, trim.outliers = T)
p3 <- partial(calif_fe_gbm, 'AveOccup', n.trees = d, trim.outliers = T)
p4 <- partial(calif_fe_gbm, 'HouseAge', n.trees = d)


grid.arrange(plotPartial(p1, ylab = '', ylim = ylims), 
             plotPartial(p2, ylab = '', ylim = ylims, xlab = 'Average Rooms'), 
             plotPartial(p3, ylab = '', ylim = ylims, xlab = 'Average Occupancy'), 
             plotPartial(p4, ylab = '', ylim = ylims, xlab = 'House Age'), 
             ncol = 2)
grid.text('Partial Predicted Log House Price', x = 0, vjust = 1, rot = 90)
```

The lack of discernible relationship between house age and predicted house price reflects its low variable importance, noting that the y-axes have been adjusted such that the features are directly comparable. For income and average rooms we observe a general positive relationship, whilst increasing average occupancy initially increases the predicted house price (although increasing the resolution could add a lot of noise between one and two occupants), before it gradually decreases.

Finally, we can also look at the partial dependence of two features simultaneously. To come full-circle with this example, we will do so for the latitude and longitude. In order to add the map border, we will extract the partial dependence information.

```{r}
#| label: fig-calif_pdp_latlong
#| fig-cap: Partial dependence plot of the location information for the gbm model fitted to the feature-engineered California housing dataset.
library(maps)
library(RColorBrewer)

latlong_pd <- plot.gbm(calif_fe_gbm, c('Longitude', 'Latitude'), return.grid = T)

# Same colour scale as before
n_breaks <- 9
my_cols <- brewer.pal(n_breaks, 'YlOrBr')[cut(latlong_pd$y, breaks = n_breaks)]

# Plot the partial dependence values
plot(latlong_pd$Longitude, latlong_pd$Latitude, pch = 15, 
     xlab = 'Longitude', ylab = 'Latitude', col = my_cols)
legend(x = 'topright', bty = 'n', title = 'Log Median House Price', cex = 0.9,
       legend = round(seq(min(latlong_pd$y), max(latlong_pd$y), length.out = n_breaks), 2),
       fill = brewer.pal(n_breaks, 'YlOrBr'))

# Add the state border
map('state', 'california', add = T, lwd = 2)
text(-123, 35, 'Pacific Ocean')
```

In @fig-calif_pdp_latlong we see that the model extrapolates beyond the state boundaries, extending the general trend observed earlier of "the more south-west, the more expensive". This completely messes up the colour scale, so to get a more realistic and useful picture let us rather only plot the partial dependence information pertaining to populated locations.

```{r}
#| label: fig-calif_pdp_latlong2
#| fig-cap: Partial dependence plot for the observed location information for the gbm model fitted to the feature-engineered California housing dataset.

library(latticeExtra)

locs <- distinct(calif, Longitude, Latitude)
p <- partial(calif_fe_gbm, c('Longitude', 'Latitude'), locs, n.trees = d)

# Same colour scale as before
n_breaks <- 9
my_cols <- colorRampPalette(brewer.pal(n_breaks, 'YlOrBr'))

pp <- plotPartial(p, col.regions = my_cols(100))

# Add the state border. plotPartial creates a lattice be default
my_map <- map_data('state', 'california')
border_layer <- latticeExtra::layer(panel.polygon(my_map, col = 'transparent'))
pp + border_layer
```

Comparing this with the original map of the data, we can see that the gbm model captured how location relates to house price.

So far we have only discussed the base version of stochastic gradient boosting, but there are several variants that use the same basic principle whilst making changes to some aspects of the algorithm. One particularly powerful, relatively recent development is Extreme Gradient Boosting (XGBoost).

### XGBoost

Since @xgboost first proposed XGBoost, it has become a highly popular choice for solving a wide range of machine learning problems, especially structured/tabular data. Conceptually, it is a gradient boosting method built from regression trees, but with several important refinements that improve statistical regularisation and computational efficiency.

To illustrate its application, we will analyse the California data one final time using the `xgboost` package. In earlier versions, it was convenient to tune XGBoost via `caret::train(method = "xgbTree")`. However, at time of writing (`r Sys.Date()`), changes in the underlying R objects created by `xgboost` v3.2+ can cause `caret` fits/tuning to fail. Therefore, we have the following options:

1. Use `xgboost`'s built-in CV grid search tuning function -- `xgb.cv()` -- which also applies **early stopping**. The only additional requirement is that the data be transformed into an `xgb.DMatrix` object. The drawback here is that hyperparameter tuning and results plotting would have to be done "manually", as opposed to using `caret`'s convenient built-in functions.

2. Use a different model training framework instead of `caret`, for example the now widely-used [tidymodels](https://workflowsets.tidymodels.org/articles/tuning-and-comparing-models.html) framework, or the more powerful (but less convenient) [mlr3](https://mlr3.mlr-org.com/) package. Although these are good options, we will not be introducing them at this stage. 

3. Install a previous version of `xgboost`. This is not generally the recommended approach for long-term usage, however, for our purposes there is little functional difference between the older version used below and the latest version. Therefore, **we will take this option**.

```{r warning=F, message=F}
install.packages("xgboost", repos = "https://p3m.dev/cran/2025-12-01")
library(xgboost)
```

The only hyperparameters we will search over are: 

- Number of trees $B$ from 5000 to 35000 in increments of 5000 
- Tree depth $d \in (3, 6, 9)$ 
- Learning rate $\lambda \in (0.005, 0.01, 0.05)$ 
- Feature subsampling per tree (similar to $m$ in the random forest, only at tree level) $\in \{0.5,  1\}$ 
- Row subsampling will be set to 0.7 

Below we see all the hyperparameters that must be specified when creating a search grid for XGBoost. Note that separate models are not run for different numbers of trees; the chain is simply extended, allowing us to use even more granular numbers of trees than this should we want to. 

Despite the relatively limited search, this will still take several hours to run, even on high performance hardware. 

```{r}
getModelInfo()$xgbTree$parameters

# calif_xgb_grid <- expand.grid(nrounds = seq(5000, 35000, 5000),  #number of trees
#                               max_depth = c(3, 6, 9),            #interaction depth
#                               eta = c(0.005, 0.01, 0.05),        #learning rate
#                               gamma = 0.001,                     #mindev
#                               colsample_bytree = c(0.5, 1),      #proportion random features per tree
#                               min_child_weight = 1,              #also controls tree depth
#                               subsample = 0.7                    #proportion of observations used per tree
#                               )
# 
# ctrl <-  trainControl(method = 'cv', number = 5, verboseIter = T)
# 
# set.seed(1)
# calif_xgb_gridsearch <- train(HousePrice ~ ., data = calif_fe_train,
#                               method = 'xgbTree',
#                               trControl = ctrl,
#                               verbose = F,
#                               tuneGrid = calif_xgb_grid)

# save(calif_xgb_gridsearch, file = 'data/xgb_calif_35k.Rdata')

load('data/xgb_calif_35k.Rdata')
```

Looking at the hyperparameter tuning CV results in @fig-calif_xgb_hps, we once again see that the deeper tree (8 leaves) is preferable across all combinations. We would ideally want to find the best tree size, since this seems to make a notable difference, especially when applying random feature selection. For this number of trees, the learning rate change does not make a substantial difference. 

```{r}
#| label: fig-calif_xgb_hps
#| fig-cap: Hyperparameter combinations results for an XGBoost model fitted to the feature-engineered California housing dataset.

plot(calif_xgb_gridsearch)
```

The best model combination (of the ones tested) is as follows:

```{r}
library(kableExtra)
kable(calif_xgb_gridsearch$bestTune)
```

The best model used tree depth equal to the highest tested value (faster learning), coupled with the lowest learning rate (slower learning). Since both these values are at the edge of the search space, one should ideally perform another search with larger and smaller values, respectively. The benefit in this instance will be negligible though, hence we will select the best model from above to predict on the test set. Likewise for finding the exact number of trees where the CV error reaches a minimum -- there is no noticeable difference in this example.

```{r message=FALSE, warning=FALSE}
options(warn = -1)
y_calif_fe <- calif_fe_test$HousePrice
calif_xgb_pred <- predict(calif_xgb_gridsearch, calif_fe_test)
calif_xgb_mse <- mean((y_calif_fe - calif_xgb_pred)^2)
```

The CV RMSE of `r round(min(calif_xgb_gridsearch$results$RMSE), 3)` was relatively close to the testing RMSE of `r round(sqrt(calif_xgb_mse), 3)` (MSE = `r round(calif_xgb_mse, 3)`), in fact slightly overestimating it. We can also determine the proportion of variation explained in the unseen data as $R^2$ = `r round((1 - sum((y_calif_fe - calif_xgb_pred)^2)/sum((y_calif_fe - mean(y_calif_fe))^2))*100, 2)`%. 

One way of visualising a model's predictions is to simply plot them against the observed values. The closer the points lie to a straight line, the better the fit. 

```{r, warning=F, message=F}
#| label: fig-calif_preds
#| fig-cap: Test set predictions from the best xgBoost model fitted to the feature-engineered California dataset plotted against the true values.

plot_data <- as.data.frame(cbind(predicted = calif_xgb_pred, observed = calif_fe_test$HousePrice))

ggplot(plot_data, aes(predicted, observed)) + geom_point(color = 'darkred', alpha = 0.5) + 
  geom_smooth(method = lm) + 
  xlab('Predicted log(price)') + ylab('Observed log(price)') + 
  theme(plot.title = element_text(color = 'darkgreen', size = 16, hjust = 0.5),
        axis.text.y = element_text(size = 12), axis.text.x = element_text(size = 12, hjust = 0.5),
        axis.title.x = element_text(size = 14), axis.title.y = element_text(size = 14)) +
  theme_bw()
```

The first thing to notice is the clear ceiling on the observed house prices, which one should ideally have already picked up in the data exploration. This maximum value of `r format(max(exp(calif$HousePrice)), scientific = F)` is suspiciously round. It turns out that there are `r sum(round(exp(calif$HousePrice)) == 500000 | round(exp(calif$HousePrice)) == 500001)` neighbourhoods where the median house price was either \$500,000 or \$500,001. Whether this is an artefact of the data due to some measurement limit or estimation, or whether it was because of some real-world bound is something the analyst should determine before providing definitive conclusions. 

For our very last example, we will return to the Titanic data.

### Example -- Titanic

Slow learning methods like gradient boosting generally tend to work better on larger datasets. Having said that, nothing prevents us from applying boosting to the Titanic data for comparison, even though it is relatively limited in size. 

The benefit of a smaller problem is of course that the training goes quicker and it generally requires fewer trees, such that one can test more hyperparameter combinations for the same time cost. Subsampling is also unlikely to yield much, if any benefit, hence we will just use the gbm method.

We will search across trees from 500 to 6000 in increments of 500; $d = 1, \ldots, 5$; and $\lambda \in (0.001, 0.005, 0.01)$.

```{r}
titanic <- read.csv('data/titanic.csv', stringsAsFactors = T)
titanic$Pclass <- as.factor(titanic$Pclass) #Class coded as integer, should be factor

set.seed(1)
train <- sample(nrow(titanic), 0.7*nrow(titanic))

titanic_train <- titanic[train,]
titanic_test <- titanic[-train,]

# set.seed(4026)
# titanic_gbm_grid <- expand.grid(n.trees = seq(500, 6000, 500),
#                                 interaction.depth = 1:5,
#                                 shrinkage = c(0.01, 0.005, 0.001),
#                                 n.minobsinnode = 1)
# 
# ctrl <-  trainControl(method = 'cv', number = 10, verboseIter = T)
# 
# titanic_gbm_gridsearch <- train(Survived ~ ., data = titanic_train,
#                                 method = 'gbm',
#                                 distribution = 'bernoulli', #For classification
#                                 trControl = ctrl,
#                                 verbose = F,
#                                 tuneGrid = titanic_gbm_grid)
# save(titanic_gbm_gridsearch, file = 'data/gbm_titanic.Rdata')
load('data/gbm_titanic.Rdata')
```

The best model used the following hyperparameters:

```{r} 
kable(titanic_gbm_gridsearch$bestTune)
```

This is a desirable result, since none of the hyperparameters are on the edge of the explored space. Using $d = 4$ and $\lambda = 0.005$, we can now find the exact number of trees where the CV deviance is minimised.

```{r}
#| label: fig-titanic_gbm_cv
#| fig-cap: Cross-validation results for the gradient boosted trees model with $\lambda=0.005$ and $d=4$ fitted to the Titanic dataset.

# gbm requires 0-1 for binary classification (without caret)
titanic_train$Survived <- as.numeric(titanic_train$Survived) - 1

set.seed(5076)
titanic_gbm <- gbm(Survived ~ ., data = titanic_train, 
                   distribution = 'bernoulli',    #For binary classification (see ?gbm)
                   n.trees = 4000, 
                   interaction.depth = 4, 
                   shrinkage = 0.005, 
                   bag.fraction = 1, 
                   cv.folds = 10, 
                   n.cores = 10, 
                   verbose = F)

d <- gbm.perf(titanic_gbm)
legend('topright', c('CV error', 'Training error'), col = c('green', 'black'), lty = 1)
```

We will use this model, with only `r d` trees, for prediction.

```{r}
y_titanic_test <- as.numeric(titanic_test$Survived) - 1
titanic_gbm_pred <- predict(titanic_gbm, newdata = titanic_test, type = 'response')
titanic_gbm_mis <- mean(y_titanic_test != round(titanic_gbm_pred))
```

Through gbm we have managed to reduce the testing misclassification rate even further, from 21.4% for the tuned random forest to `r round(titanic_gbm_mis*100, 2)`%. Exploring the confusion matrix and other classification metrics is left as an exercise. 

According to this model, passenger sex is still the most influential predictor. Age and passenger follow with similar levels of contribution. It's worth noting that the impurity and permutation variable importance measures are quite similar, contrary to the random forest model variable importance from the previous lecture, where passenger class contributed relatively more to classification accuracy during prediction than it contributed during splitting. 

```{r}
#| label: titanic-gbm-vi
#| fig-cap: Variable importance for the tuned gbm model fitted to the Titanic dataset.


set.seed(1)

x_train <- select(titanic_train, -Survived)
y_train <- titanic_train$Survived |> as.factor()

tgbm_pred_wrapper <- function(object, newdata) {
  p <- predict(object, newdata = newdata, n.trees = d, type = 'response')  # d = chosen number of trees above
  factor(as.integer(p >= 0.5), levels = levels(y_train))
}

#See vip::list_metrics() for a list of metrics

tgbm_per_imp <- vi_permute(
  object        = titanic_gbm,
  feature_names = colnames(x_train),
  train         = x_train,
  target        = y_train,
  metric        = 'accuracy',
  pred_wrapper  = tgbm_pred_wrapper,
  nsim          = 10
)

tgbm_per_p <- vip(tgbm_per_imp) +
  labs(
    title = 'Permutation Variable Importance',
    x = 'Increase in Prediction Error',
    y = NULL
  ) +
  theme_minimal()

tgbm_imp_p <- vip(titanic_gbm) +
  labs(
    title = 'Impurity Variable Importance',
    x = 'Decrease in Splitting Criterion',
    y = NULL
  ) +
  theme_minimal()

tgbm_per_p + tgbm_imp_p
```

Moving on to the partial dependence plots, we observe the expected patterns based on the classification tree's results. @fig-titanic_pdp shows that log-odds of survival were significantly increased by being female; decreased as class increases; dipped sharply after late-teens, stayed relatively stable until about 60 years of age, and then decreased again; and increased sharply around a specific fare, although there are once again outliers in fare which can be removed (this is left for the reader).

```{r}
#| label: fig-titanic_pdp
#| fig-cap: Partial dependence plots for the gbm model fitted to the Titanic dataset.
ylims = c(0,1)

p1 <- plot.gbm(titanic_gbm, 'Sex',    d, ylab = '', type = 'response', ylim = ylims)
p2 <- plot.gbm(titanic_gbm, 'Pclass', d, ylab = '', type = 'response', ylim = ylims)
p3 <- plot.gbm(titanic_gbm, 'Age',    d, ylab = '', type = 'response', ylim = ylims) 
p4 <- plot.gbm(titanic_gbm, 'Fare',   d, ylab = '', type = 'response', ylim = ylims) 

grid.arrange(p1, p2, p3, p4, ncol = 2)
grid.text('Partial Predicted Probability', x = 0, vjust = 1, rot = 90)
```

We might also be interested in the interaction between age and sex (@fig-titanic_pdp_biv) or between two continuous features such as age and fare (@fig-titanic_pdp_biv2).

```{r}
#| label: fig-titanic_pdp_biv
#| fig-cap: Bivariate partial dependence plot using sex and age for the gbm model fitted to the Titanic dataset.

#Bivariate PD plot
plot.gbm(titanic_gbm, c('Sex', 'Age'), d, ylab = '', type = 'response', ylim = ylims)
grid.text('Partial Predicted Probability', x = 0, vjust = 1, rot = 90)
```

Although the overall pattern across ages is roughly the same for males and females, we do a see a noticeably more pronounced dip specifically for males beyond age of approximately 16. 

```{r}
#| label: fig-titanic_pdp_biv2
#| fig-cap: Bivariate partial dependence plot using fare and age for the gbm model fitted to the Titanic dataset.

plot.gbm(titanic_gbm, c('Fare', 'Age'), d, type = 'response')
```

We observe little interaction between these two variables -- the marginal patterns for both mostly persist across all levels of the other.

### Final word

We end this section with one final reminder that the implementation of these supervised learning algorithms is constantly being developed, and although we mostly used `caret` for hyperparameter tuning, one could achieve the same with other libraries, e.g. [tidymodels](https://workflowsets.tidymodels.org/articles/tuning-and-comparing-models.html) or [mlr3](https://mlr3.mlr-org.com/). 

You are indeed encouraged to keep on exploring new methods in R, although one should always ensure that you have a deeply grounded understanding of the theory underpinning these methods. So continue balancing depth of theory, breadth of topic, and practical application on your future journey through supervised learning. 

Happy continued learning!

---