---
title: "STA5076Z Lecture 11 -- Bagging & Random Forests"
format: 
  html:
    fig-width: 6
    fig-height: 6
    code-fold: show
    code-summary: "Show/Hide Code"
    code-overflow: wrap
    code-tools: true
execute:
  warning: false
bibliography: references.bib
---

```{css}
/*| echo: false*/
p {
  text-align: justify;
}

table {
  width: auto !important;
  margin-left: auto;
  margin-right: auto;
}
```

The primary reason for the weak performance of decision trees on out-of-sample data is that they suffer from **high sampling variability**, especially on higher dimensional data. In other words, if we were to take different samples from the same population and fit trees to each sample, the results could look quite different, as seen in the following example.

For illustration purposes, consider the `Sonar` dataset contained in the `mlbench` package. The task is to determine whether sets of 60 sonar signals pertain to a metal cylinder/mine (M) or a roughly cylindrical rock (R). For every random 70/30 data split we fit a vanilla regression tree and track both the fitted tree and the resulting test classification accuracy.

```{r}
#| animation-hook: gifski
#| interval: 0.1
#| label: fig-variabilitree
#| fig-cap: Different trees and test accuracies resulting from different train/test splits on the Sonar dataset.

library(tree)
library(mlbench)
library(maptree) #Better tree plotting
data(Sonar)
accs <- c()

par(mfrow=c(1,2))
set.seed(4026)

for (i in 1:100){
  # Split and fit
  train <- sample(1:nrow(Sonar), 0.7*nrow(Sonar))
  my_tree <- tree(formula = Class ~ ., data = Sonar, subset = train)
  
  # Predict and evaluate
  yhat <- predict(my_tree, Sonar[-train,], type = 'class')
  y <- Sonar[-train, 'Class']
  accs <- c(accs, sum(diag(table(y, yhat)))/length(y)*100)
  
  # Plots
  draw.tree(my_tree, cex = 0.5)
  plot(1:i, accs, type = 'o', pch = 16, xlim = c(1, 100), ylim = c(50, 100),
       xlab = '', ylab = 'Classification accuracy (%)')
}
```

@fig-variabilitree shows how the fitted model and its testing accuracy can vary extensively across different random samples, with a classification accuracy range of `r round(max(accs) - min(accs), 1)`% in these 100 samples. Although we can manage the complexity -- and, therefore, the sampling variance -- to some extent through pruning, bagging and random forests offer a much more effective approach.

### Bagging decision boundaries

The benefit of the bagging procedure above a single tree is easily measured by out-of-sample performance, which in essence means that this procedure does better at capturing the true underlying relationship, $f$. This is seen by investigating the decision boundaries, as shown for simulated linearly separable data in @fig-lin_bound.

As we saw during the lecture, a bagged model is a special case of a random forest, hence we will use the `randomForest` package to fit a random forest, using all the predictors at every split. This toy example only involves two features.

```{r}
#| label: fig-lin_bound
#| fig-cap: Estimated (green) vs actual (black) decision boundaries for a single tree (left) compared to a vanilla bagged model (right).
#| fig-width: 11
#| out-width: 100%

library(randomForest)
par(mfrow = c(1, 2))

# Simulate random points
set.seed(123)
x1 <- runif(500, 0, 10)
x2 <- runif(500, 0, 10)
y <- as.factor(ifelse(x2 <= 10 - x1, 0, 1)) # With a linear decision boundary

# Tree first
mytree <- tree(y ~ x1 + x2, split = 'gini')

# Checked separately that 10 nodes is a good size (verify with cv.tree)
pruned_tree <- prune.tree(mytree, best = 10, method = 'misclass')
plot(x1, x2, col = ifelse(y == 0, 'red', 'navy'), pch = 16, xlim = c(0, 10), ylim = c(0, 10),
     xlab = expression(X[1]), ylab = '', cex.axis = 1.2, cex.lab = 1.5, asp = 1)
title(ylab = expression(X[2]), line = 1.8, cex.lab = 1.5) #ylab was being cut off

# Boundaries
abline(10, -1, lwd=3)
partition.tree(pruned_tree, add = T, lwd = 6, col = 'darkgreen', cex = 2)
x <- seq(0, 10, length = 100)
fgrid <- expand.grid(x1 = x, x2 = x)
yhat <- predict(pruned_tree, fgrid)[, 2]
contour(x, x, matrix(yhat, 100, 100), add = T,
        levels = 0.5, drawlabels = F, lwd = 4, col = 'darkgreen')

# Now bagged model (special case of random forest with m = p)
mybag <- randomForest(y ~ x1 + x2, mtry = 2)

# Plot
plot(x1, x2, col = ifelse(y == 0, 'red', 'navy'), pch = 16, xlim = c(0, 10), ylim = c(0, 10),
     xlab = expression(X[1]), ylab = '', cex.axis = 1.2, cex.lab = 1.5, asp = 1)
title(ylab = expression(X[2]), line = 1.8, cex.lab = 1.5) #ylab was being cut off

# Boundaries
abline(10, -1, lwd=3)
yhat <- predict(mybag, fgrid, type = 'prob')[, 2]
contour(x, x, matrix(yhat, 100, 100), add = T,
        levels = 0.5, drawlabels = F, lwd = 3, col = 'darkgreen')
```

Here we see that the bagged model better approximates the decision boundary. Of course, one would rather fit a linear logistic regression to such a problem, but what about scenarios where the decision boundary is especially non-linear?

```{r}
#| label: fig-circ_bound
#| fig-cap: Estimated (green) vs actual (black) decision boundaries for a single tree (left) compared to a vanilla bagged model (right).
#| fig-width: 11
#| out-width: 100%

library(randomForest)
library(plotrix) #For drawing the circle
par(mfrow = c(1, 2))

# Simulate random points
set.seed(123)
x1 <- runif(1000, 0, 10)
x2 <- runif(1000, 0, 10)
circ <- (x1 - 5)^2 + (x2 - 5)^2
y <- as.factor(ifelse(circ <= 3^2, 0, 1))

# Tree first
mytree <- tree(y ~ x1 + x2)

# Checked separately that 5 nodes is a good size (verify with cv.tree)
pruned_tree <- prune.tree(mytree, best = 5, method = 'misclass')
plot(x1, x2, col = ifelse(y == 0, 'red', 'navy'), pch = 16, xlim = c(0, 10), ylim = c(0, 10),
     xlab = expression(X[1]), ylab = '', cex.axis = 1.2, cex.lab = 1.5, asp = 1)
title(ylab = expression(X[2]), line = 1.8, cex.lab = 1.5) #ylab was being cut off

# Boundaries
draw.circle(5, 5, 3, nv = 100, border = 'black', col = NA, lwd = 4)
partition.tree(pruned_tree, add = T, lwd = 6, col = 'darkgreen', cex = 2)
x <- seq(0, 10, length = 150)
fgrid <- expand.grid(x1 = x, x2 = x)
yhat <- predict(pruned_tree, fgrid)[, 2]
contour(x, x, matrix(yhat, 150, 150), add = T,
        levels = 0.5, drawlabels = F, lwd = 4, col = 'darkgreen')

# Now bagged model (special case of random forest with m = p)
mybag <- randomForest(y ~ x1 + x2, mtry = 2)

# Plot
plot(x1, x2, col = ifelse(y == 0, 'red', 'navy'), pch = 16, xlim = c(0, 10), ylim = c(0, 10),
     xlab = expression(X[1]), ylab = '', cex.axis = 1.2, cex.lab = 1.5, asp = 1)
title(ylab = expression(X[2]), line = 1.8, cex.lab = 1.5) #ylab was being cut off

# Boundaries
draw.circle(5, 5, 3, nv = 100, border = 'black', col = NA, lwd = 4)
yhat <- predict(mybag, fgrid, type = 'prob')[, 2]
contour(x, x, matrix(yhat, 150, 150), add = T,
        levels = 0.5, drawlabels = F, lwd = 3, col = 'darkgreen')
```

@fig-circ_bound again shows that although the tree does roughly capture the overall pattern, its orthogonal splits struggle to capture the circular boundary, whereas the bagged model adjusts appropriately to some extent.

### Example -- California housing

After applying an 80/20 data split into training and testing sets, we will start by using the `randomForest` package to fit both a vanilla random forest and a bagged model to the California housing dataset, both using 250 trees. We will also keep track of the training times.

```{r}
calif <- read.csv('data/calif.csv')
calif$HousePrice <- log(calif$HousePrice)

# Train/test split
set.seed(5076)
train <- sample(1:nrow(calif), 0.8*nrow(calif))

# Bagging
bag_250_time <- system.time(
  calif_bag <- randomForest(HousePrice ~ ., data = calif, subset = train,
                            mtry = ncol(calif) - 1, #Use all features (minus response)
                            ntree = 250,
                            importance = T,         #Keep track of importance (faster without)
                            #do.trace = 25,         #Can keep track of progress if we want
                            na.action = na.exclude)
)

# Random Forest
rf_250_time <- system.time(
  calif_rf <- randomForest(HousePrice ~ ., data = calif, subset = train,
                           ntree = 250,
                           importance = T,
                           na.action = na.exclude)
)
```

From the model object we can extract much information on the fit (see `calif_rf$...`). For example, we can see that the random forest's trees contained on average `r format(round(mean(calif_rf$forest$ndbigtree)), scientific = F)` terminal nodes! We can also look at the percentage of variation in $Y$ captured by the model, i.e. the coefficient of determination ($R^2$), by just printing the model object:

```{r}
calif_bag
calif_rf
```

Verifying this calculation manually (just for the random forest model):

```{r}
y_calif <- calif[train, ]$HousePrice
yhat_calif_rf <- calif_rf$predicted
# Calculate SSR = 1 - SSE/SST
round((1 - sum((y_calif - yhat_calif_rf)^2)/sum((y_calif - mean(y_calif))^2))*100, 2)
```

Interestingly, we see that the bagged model has the higher $R^2$. Comparing their OOB error plots, we see that the bagged model also yields better (lower) errors:

```{r}
#| label: fig-calif_oobs
#| fig-cap: OOB errors for the bagged tree and random forest (default m) fitted to the California housing dataset.

plot(calif_bag$mse, type = 'l', xlab = 'Number of trees', ylab = 'OOB MSE',
     col = 'blue', lwd = 2, ylim = c(0, max(calif_rf$mse)))
lines(calif_rf$mse, col = 'darkgreen', lwd = 2, type = 's')
legend('topright', legend = c('Bagging', 'Random Forest'),
       col = c('blue', 'darkgreen'), lwd = 2)
```

Based on this, if we had to choose a model between these two, we would pick the bagged model. We will, however, apply both to the test set to compare the OOB estimates to the observed test errors. For comparison, let us add the test error for a single pruned tree as well.

```{r}
#| label: fig-calif_test
#| fig-cap: Testing errors for the bagged tree, random forest (default m), and single tree fitted to the California housing dataset, added to the OOB plots.

# Quick single tree
stopcrit <- tree.control(nobs = length(train), mindev = 0.002)
calif_tree_overfit <- tree(HousePrice ~ ., data = calif[-train, ], control = stopcrit)
set.seed(5076)
calif_cv <- cv.tree(calif_tree_overfit)
T <- calif_cv$size[which.min(calif_cv$dev)]
calif_pruned <- prune.tree(calif_tree_overfit, best = T)

# Predictions
calif_bag_pred  <- predict(calif_bag, newdata = calif[-train, ])
calif_rf_pred   <- predict(calif_rf, newdata = calif[-train, ])
calif_tree_pred <- predict(calif_pruned, newdata = calif[-train, ])

# Prediction accuracy
y_calif_test   <- calif[-train, 1]
calif_bag_mse  <- mean((y_calif_test - calif_bag_pred)^2)
calif_rf_mse   <- mean((y_calif_test - calif_rf_pred)^2)
calif_tree_mse <- mean((y_calif_test - calif_tree_pred)^2)

# Plot MSEs
plot(calif_bag$mse, type = 'l', xlab = 'Number of trees', ylab = 'MSE',
     col = 'blue', lwd = 2, ylim = c(0, calif_tree_mse*2))
lines(calif_rf$mse, col = 'darkgreen', lwd = 2)
abline(h = calif_bag_mse, col = 'blue', lty = 2, lwd = 2)
abline(h = calif_rf_mse, col = 'darkgreen', lty = 2, lwd = 2)
abline(h = calif_tree_mse, col = 'grey', lty = 2, lwd = 2)
legend('topright', legend = c('Bagging OOB', 'Bagging test', 'Random forest OOB', 'Random forest test', 'Single tree test'),
       col = c('blue', 'blue', 'darkgreen', 'darkgreen', 'grey'), lwd = 2, lty = c('solid', 'dashed', 'solid', 'dashed', 'dashed'))
```

As expected, the model with $m=8$ performed better on the test set. It is also interesting to note that for both models, the OOB error slightly *over*estimated the testing error. The bagged model's test MSE of `r round(calif_bag_mse, 2)` is approximately half of the pruned regression tree test MSE (`r round(calif_tree_mse, 2)`).

Looking at the training times (in seconds), we see that it executed fairly quickly, even for this moderately-sized dataset:

```{r}
bag_250_time
rf_250_time
```

The `ranger` package offers inter alia built-in parallelisation across cores to improve runtimes significantly. Fitting the bagged model using this package:

```{r, message=F, warning=F}
library(ranger)

ranger_time <- system.time(
  ranger_bag <- ranger(formula = HousePrice ~ ., 
                       data = calif[train, ],
                       num.trees = 250, 
                       mtry = ncol(calif) - 1)
)
```

We observe a significantly faster runtime:

```{r}
ranger_time
```

The model object also has some neat output to explore. For example, let's investigate the 100th tree in the forest:

```{r}
head(treeInfo(ranger_bag, 100))
```

Note, however, that since the trees are fitted in parallel we cannot keep track of OOB MSE as the number of trees increase; we can only extract the final OOB MSE.

We can now leverage the speed of this package to answer an important question, namely which value of $m$ is best for this example. In the analyses above we only used $m = p = 8$ and $m = \lfloor p/3 \rfloor = 2$, but perhaps another value would yield better results. To determine this, we will use `caret` to perform a **hyperparameter gridsearch** (using `ranger`) across all possible values of $m$. We can also control the tree depth, therefore we will try different size trees as well. We will again use 250 trees.

```{r message=FALSE, warning=FALSE}
library(caret)
set.seed(1)

# Create combinations of hyperparameters
# see getModelInfo()$ranger$parameters
calif_rf_grid <- expand.grid(mtry = 2:(ncol(calif) - 1),
                             splitrule = 'variance',      #Must specify. This is RSS for regression
                             min.node.size = c(1, 5, 20)) #Default for regression is 5. Controls tree size

ctrl <- trainControl(method = 'oob', verboseIter = F)     #Can set to T to track progress

# Use ranger to run all these models
calif_rf_gridsearch <- train(HousePrice ~ .,
                             data = calif,
                             subset = train,
                             method = 'ranger',
                             num.trees = 250,
                             importance = 'impurity',
                             #verbose = T,
                             trControl = ctrl,
                             tuneGrid = calif_rf_grid)    #Here is the grid
```

The hyperparameter combination that yielded the best OOB result is as follows:

```{r}
calif_rf_gridsearch$finalModel
```

According to this process, the best random forest model was one with $m =$ `r calif_rf_gridsearch$bestTune$mtry` and terminal nodes of size `r calif_rf_gridsearch$bestTune$min.node.size`. A particularly useful feature of `caret`'s `train()` function is the trivially easy plotting of the hyperparameter tuning results, allowing one to see not only which combination was best, but also how it compared to others, as well as the effect of changing certain hyperparameters. This is shown in @fig-calif_hps.

```{r}
#| label: fig-calif_hps
#| fig-cap: Hyperparameter combinations results for a random forest fitted to the California housing dataset.

plot(calif_rf_gridsearch)
```

There does not seem to be much difference between using entirely overfitted trees (minimal node size = 1) and ones with a minimal node size of 5, but increasing the node size to 20 (reducing the trees' size) yielded worse performance for all values of $m$. Although there is not much in it, we will now use the model that yielded the lowest OOB error to predict on the test set.

```{r}
calif_rfgrid_pred <- predict(calif_rf_gridsearch, newdata = calif[-train,])
mean((y_calif_test - calif_rfgrid_pred)^2)
```

Comparing this to the $m=8$ model from earlier, we actually observe a miniscule decrease in performance on the test MSE, which was `r calif_bag_mse`, although the difference is negligible. To reiterate: in practice, one would only use the test set to get one final prediction after selecting a model based on CV/OOB errors. This test set comparison was merely for illustration.

Finally, let's look at the variable importance plot. Although most packages contain their own built-in variable importance plotting functions, it is most convenient to use packages designed specifically for this purpose, like `vip`, to ensure standard presentation across different model classes. Note we measured impurity-based variable importance during training, which we can extract from the model object. We can use `vi_permute()` to calculate permutation importance (see an example of this later), or we can retrain the model and keep track of perumtation importance this time. Given the speed of ranger, we will do the latter here. 

```{r}
#| label: calif-bag-imp
#| fig-cap: Variable importance plots for the tuned random forest fitted to the California housing dataset.
#| out-width: "100%"

library(vip)
library(ggplot2)
library(patchwork) #Easy way to combine ggplots 

calif_rf_besthp <- calif_rf_gridsearch$bestTune

calif_rf_perm <- ranger(
  HousePrice ~ .,
  data = calif[train, ],
  num.trees = 250,
  mtry = calif_rf_besthp$mtry,
  min.node.size = calif_rf_besthp$min.node.size,
  splitrule = calif_rf_besthp$splitrule,
  importance = 'permutation',
  seed = 123
)

crf_imp_p <- vip(calif_rf_gridsearch) +
  labs(
    title = 'Impurity Variable Importance',
    x = 'Increase in Prediction Error',
    y = NULL
  ) +
  theme_minimal()

crf_per_p <- vip(calif_rf_perm) +
  labs(
    title = 'Permutation Variable Importance',
    x = 'Decrease in Splitting Criterion',
    y = NULL
  ) +
  theme_minimal()

crf_per_p + crf_imp_p
```

As expected, we see that median neighbourhood income yielded the greatest average reduction in RSS, during training followed by the location information. However, during permutation, removing only one of the location variables at a time did not have a major effect on prediction performance, presumably because much of the predictive power was still contained in the other location variable. 


We can now apply the same procedure to the Titanic dataset.

### Example -- Titanic

In practice one might want to jump straight to the hyperparameter tuning (after doing the necessary exploration and cleaning, of course). However, since we are still exploring and comparing these methods, we will again first see how the bagged model compares to the default random forest, with a single pruned tree again added for reference. This step also helps to determine whether the number of trees is sufficient.

Since the Titanic dataset is much smaller than the California data, we will use a 70/30 split to be more certain of our out-of-sample performance.

```{r}
#| label: fig-titanic_test
#| fig-cap: Testing errors for the bagged tree, random forest (default m), and single tree fitted to the Titanic dataset, added to the OOB error plots.

titanic <- read.csv('data/titanic.csv', stringsAsFactors = T)
titanic$Pclass <- as.factor(titanic$Pclass) #Class coded as integer, should be factor

# Train/test split
titanic <- na.omit(titanic)
set.seed(1)
train <- sample(1:nrow(titanic), 0.7*nrow(titanic))

# Bagging
titanic_bag <- randomForest(Survived ~ ., data = titanic, subset = train,
                            mtry = ncol(titanic) - 1,
                            ntree = 1000)

# Random Forest
titanic_rf <- randomForest(Survived ~ ., data = titanic, subset = train,
                           ntree = 1000)

# Quick single tree
stopcrit <- tree.control(nobs = length(train), mindev = 0.005)
titanic_tree_overfit <- tree(Survived ~ ., data = titanic, subset = train, control = stopcrit)
titanic_cv <- cv.tree(titanic_tree_overfit)
T <- titanic_cv$size[which.min(titanic_cv$dev)]
titanic_pruned <- prune.tree(titanic_tree_overfit, best = T)

# Predictions
titanic_bag_pred  <- predict(titanic_bag, newdata = titanic[-train, ])
titanic_rf_pred   <- predict(titanic_rf, newdata = titanic[-train, ])
titanic_tree_pred <- predict(titanic_pruned, newdata = titanic[-train, ], type = 'class')

# Prediction accuracy
y_titanic_test   <- titanic[-train, 1]
titanic_bag_mis  <- mean(y_titanic_test != titanic_bag_pred)
titanic_rf_mis   <- mean(y_titanic_test != titanic_rf_pred)
titanic_tree_mis <- mean(y_titanic_test != titanic_tree_pred)

# Plot errors
plot(titanic_rf$err.rate[, 'OOB'], xlab = 'Number of trees', ylab = 'Classification error',
     type = 's', col = 'darkgreen', ylim = c(0.1, titanic_tree_mis))
lines(titanic_bag$err.rate[, 'OOB'], col = 'blue', type = 's')
abline(h = titanic_bag_mis, col = 'blue', lty = 2, lwd = 2)
abline(h = titanic_rf_mis, col = 'darkgreen', lty = 2, lwd = 2)
abline(h = titanic_tree_mis, col = 'grey', lty = 2, lwd = 2)
legend('bottomright', legend = c('Bagging OOB', 'Bagging test', 'Random forest OOB', 'Random forest test', 'Single tree test'),
       col = c('blue', 'blue', 'darkgreen', 'darkgreen', 'grey'), lwd = 2, lty = c('solid', 'dashed', 'solid', 'dashed', 'dashed'))
```

For this dataset -- which is again considerably smaller than the California problem -- we see that the OOB errors underestimated the testing errors. Although the default random forest was superior, we would again like to check all possible values of $m$. We will also try two different splitting criteria: Gini index and Hellinger distance. See @hellinger for information on the latter.

```{r}
set.seed(2)

# Create combinations of hyperparameters
titanic_rf_grid <- expand.grid(mtry = 2:(ncol(titanic) - 1),
                               splitrule = c('gini', 'hellinger'),
                               min.node.size = c(1, 5, 10))

ctrl <- trainControl(method = 'oob', verboseIter = F)

# Use ranger to run all these models
titanic_rf_gridsearch <- train(Survived ~ .,
                               data = titanic,
                               subset = train,
                               method = 'ranger',
                               num.trees = 1000,
                               trControl = ctrl,
                               tuneGrid = titanic_rf_grid,
                               importance = 'impurity')

```

```{r}
#| label: fig-titanic_grid
#| fig-cap: Hyperparameter combinations results for a random forest fitted to the Titanic dataset.

plot(titanic_rf_gridsearch)
```

In @fig-titanic_grid we see that both splitting rules yield the highest accuracy when $m=4$ and minimum node size is 5, with the Gini index split being superior at this combination. Note that these models fitted to a relatively small dataset like this one can be susceptible to sampling variation. As a homework exercise, rerun the above procedure multiple times and observe the change in results.

Next, we use the model that did best in OOB performance for prediction.

```{r}
titanic_rfgrid_pred <- predict(titanic_rf_gridsearch, newdata = titanic[-train,])
titanic_rf_tuned_mis <- mean(y_titanic_test != titanic_rfgrid_pred)
```

Tuning the random forest's hyperparameters brought the misclassification rate down from `r round(titanic_rf_mis*100, 2)`% to `r round(titanic_rf_tuned_mis*100, 2)`%.

Finally, let's consider the variable importance plot. Since prediction is not the primary concern in the Titanic dataset, we will only consider impurity importance, i.e. the relative contribution by each variable to reducing the objective function during training. This was stored during training and can be extracted using `caret`'s `varImp()` function:

```{r}
#| label: fig-titanic_rf_imp
#| fig-cap: Variable importance plot for the random forest fitted to the Titanic dataset. Hyperparameter tuning has been applied to the model.

plot(varImp(titanic_rf_gridsearch))
```

Passenger sex is by far the most important factor, followed by age and fare, which were similar. Whether or not they were 3^rd^ class passengers is somewhat important (also refer back to the classification trees lecture result, where we used the entire dataset).

### Example -- Wine quality

Returning to a regression problem, specifically the wine quality dataset collected by @misc_wine_quality_186, let's see if a random forest can improve upon the $R^2$ value of 32% which we observed for the single regression tree.

```{r}
# Read in and split the data
wine <- read.csv('data/winequality-white.csv', sep = ';')

set.seed(5076)
wine_train_ind <- sample(nrow(wine), 0.8*nrow(wine))
```

Let us repeat the grid search as applied to the California dataset.

```{r}
set.seed(1)

# Create combinations of hyperparameters
wine_rf_grid <- expand.grid(mtry = 2:(ncol(wine) - 1),
                            splitrule = 'variance',      
                            min.node.size = 1:5) 

ctrl <- trainControl(method = 'oob', verboseIter = F)     

# Use ranger to run all these models
wine_rf_gridsearch <- train(quality ~ .,
                            data = wine,
                            subset = wine_train_ind,
                            method = 'ranger',
                            num.trees = 500,
                            trControl = ctrl,
                            tuneGrid = wine_rf_grid,
                            importance = 'impurity')    
```

Viewing the hyperparameter tuning results:

```{r}
#| label: fig-wine_grid
#| fig-cap: Hyperparameter combinations results for a random forest fitted to the wine dataset.

plot(wine_rf_gridsearch)
```

Regardless of the tree size, fewer features are generally better. 

What were the most important features? Based on the single tree, we except alcohol content to play a role. Here is an example of using the `vi_permute()` function to calculate the permutation performance. 

```{r}
#| label: fig-wine_rf_imp
#| fig-cap: Variable importance plots for the random forest fitted to the wine dataset. Hyperparameter tuning has been applied to the model.
#| out-width: "100%"

library(dplyr)

set.seed(1)

x_train <- select(wine[wine_train_ind, ], -quality)
y_train <- wine$quality[wine_train_ind]

vip_pred_wrapper <- function(object, newdata) {
  predict(object, newdata = newdata)
}

wine_per_imp <- vi_permute(
  object        = wine_rf_gridsearch,
  feature_names = colnames(x_train),
  train         = x_train,
  target        = y_train,
  metric        = 'rmse',
  pred_wrapper  = vip_pred_wrapper,
  nsim          = 5
)

wine_per_p <- vip(wine_per_imp) +
  labs(
    title = 'Permutation Variable Importance',
    x = 'Increase in Prediction Error',
    y = NULL
  ) +
  theme_minimal()

wine_imp_p <- vip(wine_rf_gridsearch) +
  labs(
    title = 'Impurity Variable Importance',
    x = 'Decrease in Splitting Criterion',
    y = NULL
  ) +
  theme_minimal()

wine_per_p + wine_imp_p
```

According to this model, density is much more important than the single tree suggested. 

Calculating the test RMSE:

```{r}
y <- wine[-train, 'quality']
yhat <- predict(wine_rf_gridsearch, newdata = wine[-train,])

(wine_rf_rmse <- sqrt(mean((yhat - y)^2)))
```

This is noticeably lower than the best regression tree's RMSE of 0.71. The random forest also yields a much improved coefficient of determination:

```{r}
(1 - sum((y - yhat)^2)/sum((y - mean(y))^2))*100
```

---